Artificial inteligence-assisted analysis of CT abnormalities during COVID-19 recovery

### 2023-09-12 report/markdown/report.Rmd
Figures, tables and report RMD

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPDF", 
                      dpi = 600)

@@

flextable::flextable(tables$cohort) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2:5, width = 3.0, unit = 'cm') %>% 
  width(6, width = 2.7, unit = 'cm') %>% 
  width(7, width = 2.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('BMI: body mass index; COPD: chronic obstructive pulmonary disease; CKD: chronic kidney disease'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 6:6, 
           value = as_paragraph('Comparison between COVID-19 severity groups. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic, numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected for multiple testing with the false discovery rate method'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(4, 1, 
           value = as_paragraph('young adult: 18 - 30 years; middle aged: 31 - 65 years; elderly: 66 years and more'), 
           part = 'body', 
           ref_symbols = 'c') %>% 
  footnote(8, 1, 
           value = as_paragraph('normal: BMI < 25 kg/m\u00B2; overweight: BMI 25 - 30 kg/m\u00B2; obesity: BMI > 30 kg/m\u00B2'), 
           part = 'body', 
           ref_symbols = 'd') %>% 
  theme_vanilla

@@

## Chest computed tomography paramaters and findings

@@

flextable::flextable(tables$cohort_ct) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3:6, width = 3.8, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity; CTSS: CT severity score; AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

@@

flextable::flextable(tables$cohort_lft) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3:6, width = 3.8, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second; DLCO: diffusion capacity for CO; FEV1:FVC: FEV1 to FVC ratio'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

@@

flextable::flextable(tables$ai_cutoffs) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 4.8, unit = 'cm') %>% 
  width(4:8, width = 2.5, unit = 'cm') %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3, 
           value = as_paragraph('AUC: area under the curve, receiver - operating characteristic with 95% confidence interval.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 7, 
           value = as_paragraph('J: Youden J statistic'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 8, 
           value = as_paragraph('PPV: positive prediction value'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
   footnote(1, 9, 
           value = as_paragraph('NPV: negative prediction value'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

@@

@@

flextable::flextable(tables$ai_cutoffs_severity) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4:7, width = 5.2, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 3, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

flextable::flextable(tables$ai_cutoffs_fup) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4:7, width = 5.2, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 3, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

@@

flextable::flextable(tables$ild_cutoff) %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2, width = 4.8, unit = 'cm') %>% 
  width(3:8, width = 2.5, unit = 'cm') %>% 
  footnote(1, 1, 
           value = as_paragraph('AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('AUC: area under the curve, receiver - operating characteristic with 95% confidence interval.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 6, 
           value = as_paragraph('J: Youden J statistic'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 7, 
           value = as_paragraph('PPV: positive prediction value'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
   footnote(1, 8, 
           value = as_paragraph('NPV: negative prediction value'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

@@

flextable::flextable(tables$ild_cutoff_severity) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2:5, width = 5, unit = 'cm') %>% 
  footnote(1, 2, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(tables$ild_cutoff_fup) %>% 
  width(1, width = 2.7, unit = 'cm') %>% 
  width(2:5, width = 5, unit = 'cm') %>% 
  footnote(1, 2, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

@@

## Development of multi-paramater Random Forest models of lung function variables and abnormalities

@@

```{r tab-rf-tuning, tab.cap = 'Optimal values of the mtry, splitrule and min.node.size paramaters for Random Forest models of lung function abnormalities and paramaters found by cross-validation tuning.'}

flextable::flextable(tables$rf_tuning) %>% 
  width(1, width = 2.5, unit = 'cm') %>% 
  width(2, width = 2.7, unit = 'cm') %>% 
  width(3, width = 3, unit = 'cm') %>% 
  width(4:6, width = 2.7, unit = 'cm') %>% 
  merge_v(1:2) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for CO; FVC: forced vital capacity; FEV1: forced expiratory volume in 1 second'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

## Performance of Random Forest models at prediction of lung function

@@

```{r tab-bin-models, tab.cap = 'Performance of Random Forest classifiers at prediction of lung function testing (LFT) abnormalities in the training dataset and cross-validation (CV). CV results for any LFT abnormalities and diffusion capacity for CO (DLCO) < 80% reference are presented. The entire table is available in a supplementary Excel file.'}

>>>>

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

@@

flextable::flextable(tables$cohort) %>% 
  set_widths(c(3, rep(3.5, 4), 2.7, 2.5)) %>% 
  footnote(1, 1, 
           value = as_paragraph('BMI: body mass index; COPD: chronic obstructive pulmonary disease; CKD: chronic kidney disease'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 6:6, 
           value = as_paragraph('Comparison between COVID-19 severity groups. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic, numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected for multiple testing with the false discovery rate method'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(4, 1, 
           value = as_paragraph('young adult: 18 - 30 years; middle aged: 31 - 65 years; elderly: 66 years and more'), 
           part = 'body', 
           ref_symbols = 'c') %>% 
  footnote(8, 1, 
           value = as_paragraph('normal: BMI < 25 kg/m\u00B2; overweight: BMI 25 - 30 kg/m\u00B2; obesity: BMI > 30 kg/m\u00B2'), 
           part = 'body', 
           ref_symbols = 'd') %>% 
  theme_vanilla

@@

## Chest computed tomography abnormalities

```{r fig-ct-finding-course, fig.width = figur::convert(figures$ct_finding_course, to = 'in')$w, fig.height = figur::convert(figures$ct_finding_course, to = 'in')$h, fig.cap = 'Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

figures$ct_finding_course$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ct-finding-course). Frequency of computed tomography abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Frequencies of any chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulations and consolidations at the consecutive follow-ups after COVID-19 in the CovILD study patients stratified by the severity of acute COVID-19 were presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars.
</p>

```{r fig-ct-numeric-course, fig.width = figur::convert(figures$ct_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$ct_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

figures$ct_numeric_course$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ct-numeric-course). Time course of numeric computed tomography readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Severity of chest computed tomography (CT) abnormalities was assessed by a human-determined CT severity score (CTSS), and AI-determined percentages of the lung tissue with opacity and high opacity. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points.
</p>

@@

flextable::flextable(tables$cohort_ct) %>% 
  set_widths(c(2.7, 2.7, rep(3.8, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity; CTSS: CT severity score; AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

```{r fig-lft-finding-course, fig.width = figur::convert(figures$lft_finding_course, to = 'in')$w, fig.height = figur::convert(figures$lft_finding_course, to = 'in')$h, fig.cap = 'Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

figures$lft_finding_course$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-lft-finding-course). Frequency of lung function testing abnormalities at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Frequency of any lung function testing (LFT) abnormalities, as well as percentage of patients with DLCO, FVC and FEV1 values below 80% of the reference values at the consecutive follow-ups after COVID-19 in the CovILD cohort stratified by the severity of acute COVID-19 was presented in stacked bar plots. Numbers of complete observations at the follow-up examinations are displayed above the data bars.
<br>
LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second. 
</p>

```{r fig-lft-numeric-course, fig.width = figur::convert(figures$lft_numeric_course, to = 'in')$w, fig.height = figur::convert(figures$lft_numeric_course, to = 'in')$h, fig.cap = 'Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.'}

figures$lft_numeric_course$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-lft-numeric-course). Time course of numeric lung function testing readouts at the consecutive follow-ups in the CovILD participants stratified by severity of acute COVID-19.</b>
<br>
Lung function was assessed by measurement of DLCO, FVC and FEV1 expressed as percentage of the age-, weight- and sex-specific reference. Time courses of these parameters are visualized in box plots. Boxes represent medians with interquartile ranges, whiskers span over 150% of the interquartile range. Single observations are depicted as points. Numbers of complete observations at the follow-up examinations are displayed above the data points.
<br>
LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.
</p>

@@

flextable::flextable(tables$cohort_lft) %>% 
  set_widths(c(2.7, 2.7, rep(3.8, 4))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('LFT: lung function testing; FVC: forced vital capacity; FEV1: forced expiratory volume in one second; DLCO: diffusion capacity for carbon monoxide; FEV1:FVC: FEV1 to FVC ratio'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

```{r fig-ai-opacity-roc, fig.width = figur::convert(figures$ai_opacity_roc, to = 'in')$w, fig.height = figur::convert(figures$ai_opacity_roc, to = 'in')$h, fig.cap = 'Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung opacity.'}

figures$ai_opacity_roc$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ai-opacity-roc). Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung opacity.</b>
<br>
The optimal cutoffs of the artificial intelligence (AI) determined lung opacity for detection of the human-recognizable chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulation, and consolidations were found by maximizing the Youden's J statistic (i.e. sensitivity + specificity - 1). Results of the optimal opacity cutoff finding are presented as receiver-operating characteristic (ROC) curves. The optimal opacity cutoffs are labeled with points and the corresponding opacity values. Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plots. 
</p>

```{r fig-ai-high-opacity-roc, fig.width = figur::convert(figures$ai_high_opacity_roc, to = 'in')$w, fig.height = figur::convert(figures$ai_high_opacity_roc, to = 'in')$h, fig.cap = 'Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung high opacity.'}

figures$ai_high_opacity_roc$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ai-high-opacity-roc). Detection of human-recognizable abnormalities of chest computed tomography by artificial intelligence-measured lung high opacity.</b>
<br>
The optimal cutoffs of the artificial intelligence (AI) determined lung high opacity for detection of the human-recognizable chest computed tomography (CT) abnormalities, ground glass opacities (GGO), reticulation, and consolidations were found by maximizing the Youden's J statistic (i.e. sensitivity + specificity - 1). Results of the optimal high opacity cutoff finding are presented as receiver-operating characteristic (ROC) curves. The optimal high opacity cutoffs are labeled with points and the corresponding opacity values. Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plots. 
</p>

@@

flextable::flextable(tables$ai_cutoffs) %>% 
  set_widths(c(2.7, 3, 4.8, rep(2.5, 5))) %>% 
  merge_v(1) %>% 
  footnote(1, 2, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 3, 
           value = as_paragraph('AUC: area under the curve, receiver - operating characteristic with 95% confidence interval.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 7, 
           value = as_paragraph("J: Youden's J statistic"), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 8, 
           value = as_paragraph('PPV: positive prediction value'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
   footnote(1, 9, 
           value = as_paragraph('NPV: negative prediction value'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

@@

```{r fig-ai-opacity-severity, fig.width = figur::convert(figures$ai_opacity_severity, to = 'in')$w, fig.height = figur::convert(figures$ai_opacity_severity, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.'}

figures$ai_opacity_severity$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ai-opacity-severity). Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.</b>
<br>
The optimal cutoffs of artificial intelligence (AI) determined lung opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-opacity-roc). Concordance between the CT findings identified by lung opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and the CovILD study participants stratified by the severity of acute COVID-19 was assessed by Cohen's $\kappa$ inter-rated reliability statistic. $\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots. Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
<br>
CT: computed tomography; GGO: ground glass opacity.
</p>

```{r fig-ai-opacity-fup, fig.width = figur::convert(figures$ai_opacity_fup, to = 'in')$w, fig.height = figur::convert(figures$ai_opacity_fup, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.'}

figures$ai_opacity_fup$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ai-opacity-fup). Performance of artificial intelligence-determined lung opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.</b>
<br>
The optimal cutoffs of artificial intelligence (AI) determined lung opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-opacity-roc). Concordance between the CT findings identified by lung opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and at consecutive post-COVID-19 follow-up examinations was assessed by Cohen's $\kappa$ inter-rated reliability statistic. $\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots. Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
<br>
CT: computed tomography; GGO: ground glass opacity.
</p>

```{r fig-ai-high-opacity-severity, fig.width = figur::convert(figures$ai_high_opacity_severity, to = 'in')$w, fig.height = figur::convert(figures$ai_high_opacity_severity, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.'}

figures$ai_high_opacity_severity$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ai-high-opacity-severity). Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities in patients stratified by severity of acute COVID-19.</b>
<br>
The optimal cutoffs of artificial intelligence (AI) determined lung high opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-high-opacity-roc). Concordance between the CT findings identified by lung high opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and the CovILD study participants stratified by the severity of acute COVID-19 was assessed by Cohen's $\kappa$ inter-rated reliability statistic. $\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots. Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
<br>
CT: computed tomography; GGO: ground glass opacity.
</p>

```{r fig-ai-high-opacity-fup, fig.width = figur::convert(figures$ai_high_opacity_fup, to = 'in')$w, fig.height = figur::convert(figures$ai_high_opacity_fup, to = 'in')$h, fig.cap = 'Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.'}

figures$ai_high_opacity_fup$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ai-high-opacity-fup). Performance of artificial intelligence-determined lung high opacity at detection of chest computed tomography abnormalities at post-COVID-19 follow-up evaluations.</b>
<br>
The optimal cutoffs of artificial intelligence (AI) determined lung high opacity for detection of human-recognizable lung computed tomography (CT) abnormalities were computed as presented in Figure \@ref(fig:fig-ai-high-opacity-roc). Concordance between the CT findings identified by lung high opacity using the optimal cutoffs and the CT abnormalities determined by the radiologist in the entire observation data set ('cohort') and at consecutive post-COVID-19 follow-up examinations was assessed by Cohen's $\kappa$ inter-rated reliability statistic. $\kappa$ values with their 95% confidence intervals obtained by bootstrapping with B = 2000 iterations are presented in Forest plots. Numbers of complete observations ('total') and observations with the given CT abnormality ('events') are indicated in the Y axes.
<br>
CT: computed tomography; GGO: ground glass opacity.
</p>

@@

flextable::flextable(tables$ai_cutoffs_severity) %>% 
  set_widths(c(2.7, 3, 3, rep(5.2, 4))) %>% 
  merge_v(1:2) %>% 
  footnote(1, 3, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

flextable::flextable(tables$ai_cutoffs_fup) %>% 
  set_widths(c(2.7, 3, 3, rep(5.2, 4))) %>% 
  merge_v(1:2) %>% 
  footnote(1, 3, 
           value = as_paragraph('CT: chest computed tomography; GGO: ground glass opacity'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r fig-ctss-ild-cutoff, fig.width = figur::convert(figures$ctss_ild_cutoff, to = 'in')$w, fig.height = figur::convert(figures$ctss_ild_cutoff, to = 'in')$h, fig.cap = 'Detection of interstitial lung disease defined by opacity >5% of the lung by a human-determined computed tomography severity score.'}

figures$ctss_ild_cutoff$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ctss-ild-cutoff). Detection of interstitial lung disease defined by opacity >5% of the lung by a human-determined computed tomography severity score.</b>
<br>
(A) The optimal cutoff of the radiologist's computed tomography severity score (CTSS) for diagnosis of interstitial lung disease (ILD) defined as opacity > 5% of the lung was found by maximizing the Youden's J statistic (i.e sensitivity +  specificity - 1). Results of the optimal CTSS cutoff finding are presented as a receiver-operating characteristic (ROC) curve. The optimal CTSS cutoff is labeled with a point and the corresponding CTSS value. Sensitivity, specificity, Youden's J and area under the ROC curve (AUC) are displayed in the plot.
<br>
(B) Concordance between the ILD diagnosis with CTSS > 10 points and ILD defined as opacity > 5% of the lung in the whole observation set ('cohort') and in observations stratified by the acute disease severity or post-COVID-19 follow-up was assessed by Cohen's $\kappa$ inter-rater reliability statistic. $\kappa$ values with 95% confidence intervals obtained by bootstrap with B = 2000 iterations are presented in Forest plots. Numbers of complete observations ('total') and cases of ILD defined by >5% lung opacity ('events') are indicated in the Y axes.
</p>

@@

flextable::flextable(tables$ild_cutoff) %>% 
  set_widths(c(3, 4.8, rep(2.6, 6))) %>% 
  footnote(1, 1, 
           value = as_paragraph('AI: artificial intelligence'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('AUC: area under the curve, receiver - operating characteristic with 95% confidence interval.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 6, 
           value = as_paragraph('J: Youden J statistic'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  footnote(1, 7, 
           value = as_paragraph('PPV: positive prediction value'), 
           part = 'header', 
           ref_symbols = 'd') %>% 
   footnote(1, 8, 
           value = as_paragraph('NPV: negative prediction value'), 
           part = 'header', 
           ref_symbols = 'e') %>% 
  theme_vanilla

@@

flextable::flextable(tables$ild_cutoff_severity) %>% 
  set_widths(c(2.7, rep(5, 4))) %>% 
  footnote(1, 2, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(tables$ild_cutoff_fup) %>% 
  set_widths(c(2.7, rep(5, 4))) %>% 
  footnote(1, 2, 
           value = as_paragraph('\u03BA: Cohen \u03BA inter-rater reliability statistic'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

```{r fig-ctss-ai-relation, fig.width = figur::convert(figures$ctss_ai_relation, to = 'in')$w, fig.height = figur::convert(figures$ctss_ai_relation, to = 'in')$h, fig.cap = "Mathematical relationship between the artificial intelligence-determined lung opacity and high opacity, and the radiologist's computed tomography severity score."}

figures$ctss_ai_relation$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-ctss-ai-relation). Mathematical relationship between the artificial intelligence-determined lung opacity and high opacity, and the radiologist's computed tomography severity score.</b>
<br>
Mathematical relationships (A) between the artificial intelligence's lung opacity and radiologist's CTSS (computed tomography severity score), and (B) between the artificial intelligence's high lung opacity and radiologist's CTSS were explored by ordinary least squares 4^th^ order polynomial regression, robust 4^th^ order polynomial regression or generalized additive modeling (GAM) with thin plate splines. The model fit quality was assessed by the pseudo-R^2^ metric of explained variance and median absolute error (MedAE). The relationships are presented in scatter plots with single observations visualized as points. The model fits with their 2 $\times$ standard error regions are represented by blue lines with gray ribbons. The R^2^ and MedAE values along with numbers of complete observations are displayed in the plot captions.
</p>

@@

## Development of multi-paramater machine learning models of lung function variables and abnormalities

@@

```{r fig-dlco-red-tuning, fig.width = figur::convert(figures$dlco_red_tuning, to = 'in')$w, fig.height = figur::convert(figures$dlco_red_tuning, to = 'in')$h, fig.cap = 'Tuning of machine learning models of insufficient diffusion capacity for carbon monoxide.'}

figures$dlco_red_tuning$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-dlco-red-tuning). Tuning of machine learning models of insufficient diffusion capacity for carbon monoxide.</b>
<br>
Insufficient diffusion capacity for carbon monoxide (DLCO) was defined as values < 80% of the age-, sex-, and weight-specific reference. Probability of insufficient DLCO was modeled as a function of 
`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits. 
The models were built with the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial). The optimal sets of algorithm's parameters such as number of trees for GBM, number of randomly selected variables for random forests, size of the hidden layer for neuronal networks or cost penalty value for SVM were found by maximizing the sum of sensitivity and specificity, i.e. Youden's J criterion, in 10-repeats 10-fold cross-validation - in a process called 'model tuning'. Values of the J statistic for various combination of the algorithm's parameters are presented in plots. The optimal parameter combinations corresponding to the maximal J value are displayed in the plot captions.
</p>

```{r fig-dlco-red-resamples, fig.width = figur::convert(figures$dlco_red_resamples, to = 'in')$w, fig.height = figur::convert(figures$dlco_red_resamples, to = 'in')$h, fig.cap = 'Performance of machine learning models of insufficient diffusion capacity for carbon monoxide in cross-validation folds.'}

figures$dlco_red_resamples$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-dlco-red-resamples). Performance of machine learning models of insufficient diffusion capacity for carbon monoxide in cross-validation folds.</b>
<br>
Machine learning models of insufficient diffusion capacity for carbon monoxide (DLCO < 80% reference value) employing the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial) were developed as presented in Figure \@ref(fig:fig-dlco-red-tuning). Their performance at detection of the reduced DLCO in 10-repeats 10-fold cross-validation was measured by Youden's J statistic, i.e. the sum of sensitivity and specificity. Sorted J values are presented in plots. Orange dashed lines visualize J values expected for a meaningless model.
</p>

```{r fig-dlco-tuning, fig.width = figur::convert(figures$dlco_tuning, to = 'in')$w, fig.height = figur::convert(figures$dlco_tuning, to = 'in')$h, fig.cap = 'Tuning of machine learning regression models of diffusion capacity for carbon monoxide.'}

figures$dlco_tuning$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-dlco-tuning). Tuning of machine learning regression models of diffusion capacity for carbon monoxide.</b>
<br>
Diffusion capacity for carbon monoxide (DLCO) expressed as percentage of he age-, sex-, and weight-specific reference was modeled as as a function of 
`r length(lft_globals$baseline_variables) + length(lft_globals$ct_variables)` explanatory variables concerning demographic and clinical history, course of acute COVID-19, follow-up time as well as lung computed tomography results obtained at consecutive follow-up visits. The models were built with the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial). The optimal sets of algorithm's parameters such as number of trees for GBM, number of randomly selected variables for random forests, size of the hidden layer for neuronal networks or cost penalty value for SVM were found by minimizing mean absolute error (MAE) in 10-repeats 10-fold cross-validation - in a process called 'model tuning'. Values of the MAE statistic for various combination of the algorithm's parameters are presented in plots. The optimal parameter combinations corresponding to the minimal MAE are displayed in the plot captions.
</p>

```{r fig-dlco-resamples, fig.width = figur::convert(figures$dlco_resamples, to = 'in')$w, fig.height = figur::convert(figures$dlco_resamples, to = 'in')$h, fig.cap = 'Performance of machine learning regression models of diffusion capacity for carbon monoxide in cross-validation folds.'}

figures$dlco_resamples$plot

<p class = "legend">
<b>Figure \@ref(fig:fig-dlco-resamples). Performance of machine learning regression models of diffusion capacity for carbon monoxide in cross-validation folds.</b>
<br>
Machine learning models of diffusion capacity for carbon monoxide (DLCO, percentage of the reference value) employing the following algorithms: gradient boosted machines (GBM), random forest, neural network with a single hidden layer, and support vector machines with radial kernel (SVM radial) were developed as presented in Figure \@ref(fig:fig-dlco-tuning). Their performance at detection of the reduced DLCO in 10-repeats 10-fold cross-validation was measured by mean absolute error (MAE). Sorted MAE values are presented in plots.
</p>

```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm paramaters by cross-validation-mediated tuning.'}

flextable::flextable(tables$tuning) %>% 
  set_widths(c(3, 6, 6)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion lung capacity for carbon monoxide; FCV: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
          ref_symbols = 'a', 
          part = 'header') %>% 
  theme_vanilla

@@

## Performance of machine learning models at prediction of lung function testing abnormalities

@@

```{r fig-bin-performance, fig.width = figur::convert(figures$bin_performance, to = 'in')$w, fig.height = figur::convert(figures$bin_performance, to = 'in')$h, fig.cap = 'Performance of machine learning models at prediction of abnormalities of lung function testing in the training data set and repeated cross-validation.'}

figures$bin_performance$plot
